### 2021-08-26 plos_latex_template.tex
RJ: Andrea's edits to organizing team

Conferences are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions. The larger the conference, the larger the opportunities to network and learn from your cohort. However, conferences can become an exclusive space for privileged groups (e.g., white, male, from a rich country, English-native speaker, with no physical disabilities)  \cite{arendDisparityConferenceRegistration2019, timperleyHeMoanaPukepuke2020, gewinWhatScientistsShould2019, brownAbleismAcademiaWhere2018}.

%andrea will write a paragraph about why diversity and the systemic inequalities
While there has been some action to address systematic inequalities, there is still a lot of room for improvement.
% Rocío: Somehow this part below should be linked to the systemic inequalities part
Each one of us can have different points of view and experiences regarding inequality, depending on where we come from.
If you are reading this, you may be here because you already have some experience in Diversity, Inclusion, and Accessibility practices. Or you may want to improve diversity in your conference but do not know how or where to start. This article does not intend to be a comprehensive review of this topic --this would be impossible-- but will give some ideas pointing in key directions.   %Rocío: And finish with something more specific to connect with the next paragraph.

This article suggests rules to pivot traditional conferences towards inclusiveness and diversity and to welcome minoritized groups. It is directed as people who are part of a stable meetings commitee that oversees the site/location selection process or that coordinates with the local organizers (e.g. R Foundation Conference Committee, Ecological Society of America Meetings Committee). It is also directed at the local/virtual organizers who desire to make an inclusive conference starting at the planning stage.

These tips stem from the authors' experience of organizing useR! 2021, a virtual and global conference for users and developers of the R programming language \cite{r_core_team_2021}. We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, truly inclusive, accessible, and welcoming experience for everyone. Most of the authors also have experience of organizing other regional and national academic conferences and events in communities of practice such as R-Ladies and The Carpentries. Here, we share the lessons learned within the past year of organizing this global useR! conference, summarized as 10 simple rules towards an inclusive conference.

@@

Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity,  geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few. Human diversity should be celebrated and respected in every way. Nonetheless, there are implicit hierarchies along these axes, and some statuses (e.g., cisgender, white, male, from the US or Europe) hold the privilege of being the default settings for which all systems --including conferences-- are consciously and unconsciously built. As a result, most people in the world fall into some category where they are part of a marginalized group --as opposed to a privileged one. The first step towards a diverse and inclusive conference is to recognize that people are diverse, and that some groups face discrimination (some more than others) and might be underrepresented in your community and event \cite{timperleyHeMoanaPukepuke2020}.

Think about the groups that are usually unsupported or discriminated against in your community or event, and about the meaning of "diverse" in your context. Imagine the result you would like to see if you succeeded in organizing a diverse conference according to your vision. Would this translate into an even gender distribution in your speakers? Would it be representation of all races -especially Black people-- among the head organizers, speakers, and attendees? Would it be having LGBTQ+ friendly-spaces or community participation from key geographic regions? This vision should guide and help hold the organizing team accountable along the way.

@@

@@

%So take this seriously, think about the reasons why you want to create a more diverse conference, and be open to learn. Do not treat diversity as a checklist, do not use people from minoritized groups for image purposes (tokenization), and do not act as a savior but as a genuine collaborator. %andrea: some version of this might come in a box.

@@

\section{Have a strong online component of the conference} 
\label{rule_online}
In-person interaction is priceless; however, it is more expensive for some, and even unattainable for others. This inaccessibility is particularly true for global conferences that usually take place in high-income countries, making it financially demanding for international participants and often impossible to attend due to immigration requirements \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019}. Online conferences are more inclusive: they do not need a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities. This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinsonJournalMedicine20202021}. Furthermore, an online format is more environmentally friendly since it eliminates travel-related emissions \cite{sarabipourChangingScientificMeetings2021,ninerBetterWhomLeveling2021, gattrellComparisonCarbonCosts2021}.

@@

Gathering a diverse team will only work if there is real inclusion. Disabled people often say: "Nothing about us without us"; the same holds for other dimensions of diversity. This means that the actual life experiences, expertise, and insights from people in marginalized groups are not replaceable with good intentions from people outside these groups. A truly inclusive and welcoming space is one in which everyone in the team is invited and allowed to bring their experience to bear. So, if you plan to organize something related to any community, region, or specific dimension of diversity, involve people from these groups. Having diverse people in decision-making positions will affect positively all the other aspects of your conference because all the processes will benefit from their input, expertise, and distinct perspectives \cit{hong_groups_2004}. On the other hand, seeing diverse teams is important to create a welcoming space because seeing people with similar life experiences occupy public spaces, positions of power, and breaking negative stereotypes (i.e., representation) matters (see Rule \ref{rule_inclusion}). %Representation is an important aspect of diversity and inclusion: seeing people like you, from your country, that speak your language is one of the best way to feel you belong

Creating and maintaining such a team and space may seem more challenging than working in homogeneous teams, but the positive outcomes far outweigh any inconveniences and the team will grow, learn, and embrace more inclusive practices \cite{hongGroupsDiverseProblem2004}.

%How to work: 
Most advice when working with a diverse team should apply to every team, but is especially true when working with people from marginalized groups. First, delegate. One would be surprised by how often conference leads tend to do everything themselves instead of reaching out to the right people for the right task -- likely because the latter involves careful planning and communication well ahead of time or because they naively overestimate their expertise in areas they lack practical experience in.

Don't expect self-nomination and voting to work as mechanisms to counteract systemic inequalities. Nominate directly and offer the positions of power, even if this means stepping down. Offer mentorship and guidance if you encounter doubts, cases of impostor syndrome, doubts about the use of English in communication \ref{rule_language}. Break the expectations about lonley leadership and create smaller, co-led groups.

In addition to exclude derogatory or discriminatory language (which would constitute a violation of the code of conduct \ref{rule_inclusion} but also, should never happen), make the effort to teach yourself the best ways to communicate to account for every culture and situation.

Finally, splitting the workload and responsibilities should not be done by putting care-taking labors --community building, meeting organization and note taking, conversations with potential partners-- on the hands of women and other minoritized groups .

Take care of your team!

@@

@@

Most importantly, accessibility practices are not afterthoughts that can be dealt with at the last minute. 
They require time and early decision-making \cite{irishIncreasingParticipationUsing2020}. 
Conversely, inaccessible decisions are hard to course-correct, e.g., when finding out too late that a venue is inaccessible for wheelchairs in an in-person conference. 
On the other hand, it is likely that financial constraints may not allow you to make your conference completely accessible and there will be choices to make (e.g. choosing between captions and interpreters). 
Welcoming people that are usually disabled by standard practices in conferences into the organizing team would allow them to take part in the decisions from the beginning, spot right away the inaccessible practices that need to improve (see \textbf{Rule \ref{rule_organizing_team}}), and help make the right choices.

@@

\section{Don't let perfect be the enemy of good}
 While conference teams should have a bold vision and clear targets consensus about where to shine and where to compromise is equally important. As a team, this is a challenging journey. It means letting go of some ideals, or committing time and money into something that one does not find particularly useful or important.
%Yani: ughhhh... this doesn't sound good. I agree more with let go some ideal, for example, when I compromise in not have caption for tutorials, because we don't have money. Have caption in the other part of the conf was better because reach a more wide audience of the conference. I will try to show the compromise from the point of view of all the things we know we still have to do about inclusion, but we cannot do because we don't have money or the knowledge or the capacity at this time. This is also mention in some of the other rules. I don't remember if it is the one with the goals and indicators.

Remember, aspirations can be local and narrow and as a consequence shy away valuable contributors. A late bus is almost certainly better than no bus. 
% Yani: I don't like this phrase. It is like we have something of a lower quality than the original idea, when I truly think we have a high quality conference because was inclusive and diverse. and science say that a diverse team get better results :)
A broader pool of contributors does not only allow organizers to add more and work on multiple aspects at a different pace, it also helps to split workload and responsibilities to avoid burning out over the marathon that organizing a conference is.
%Yani: We have a broader team because we want a more diverse pool of view and experience, this is a good phrase but sound like we people were an object. Perhaps we can rephrase and add to the rule that talk about the composition of the team.%andrea: i am taking split workload and responsibilities up, but they were already there.

Without allowing for compromises, workload will naturally channel to those who are committed the most and already carry the largest share of the load. Hence compromises are no sign of giving in but are crucial to delegation, building self-responsible teams and keep a good spirit in the long run.

%\section{Rule X: Have a bold vision but be ready to compromise when necessary}
This rule is a complement to all the other rules. It is important to have a bold vision and clear targets. Some parts might be defined as baseline assumptions that are not up for discussion. However, there will always be compromises. Ideal solutions that meet all our expectations are often not possible. Don't let the perfect be the enemy of good. As a team, this is a challenging journey. It means letting go of some ideals, or committing time and money into something that one does not find particularly useful or important. We organized useR! during a global pandemic. Very early on, we tried to recognize that our resources were limited (both in time of the organizing team but also in money). We had to let go of certain aspects that were important to many of us. For an inclusive conference, we need to avoid giving in because considering all aspects seems impossible and burning ourselves and the team out because considering all aspects is actually impossible.

@@

Since accessibility and inclusiveness are not as widespread as they should be, the process might seem exacting.
But this is where the power of communities and individuals reside. 
If more conferences and domains apply these rules, the process will get more streamlined, straightforward, and mainstream to adapt with minimal overhead. 
And you will make a change towards healthier, stronger, and more inclusive communities, empowering historically marginalized people and learning together, thereby increasing the overall quality of your conferences.

We engaged in most of the practices mentioned here, and learned others along the way, so we share them here as part of our learning process. 
We organized useR! during a global pandemic  
and, as a team, this was a challenging journey. 
Very early on, we tried to recognize that our resources were limited (both in time of the organizing team and financially). 
We had to let go of certain aspects that were important to many of us. From that experience, this is our last message: 
Your conference may not become perfectly inclusive and accessible. 
Don't let perfect be the enemy of good.

>>>>

Conferences are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions. The larger the conference, the larger the opportunities to network and learn from your cohort. However, conferences can become discriminating spaces, in which members of some specific privileged groups (e.g., white, male, from a rich country, English-native speaker, with no physical disabilities) reproduce the systematic inequalities that occur in academia and society \cite{arendDisparityConferenceRegistration2019, timperleyHeMoanaPukepuke2020, gewinWhatScientistsShould2019, brownAbleismAcademiaWhere2018}.

While there has been some action to address systematic inequalities [where], there is still a lot of room for improvement.
% Rocío: Somehow this part below should be linked to the systemic inequalities part
Each one of us can have different points of view and experiences regarding inequality, depending on where we come from.
If you are reading this, you may be here because you already have some experience in Diversity, Inclusion, and Accessibility practices. Or you may want to improve diversity in your conference but do not know how or where to start. This article does not intend to be a comprehensive review of this topic --this would be impossible-- but will give some ideas pointing in key directions.   %Rocío: And finish with something more specific to connect with the next paragraph.

This article suggests rules to pivot traditional conferences towards inclusiveness and diversity. It is directed as people who are part of a stable meetings committee that oversees the site/location selection process or that coordinates with the local organizers (e.g. R Foundation Conference Committee, Ecological Society of America Meetings Committee). It is also directed at the local/virtual organizers who desire to make an inclusive conference starting at the planning stage.

These tips stem from the authors' experience of organizing useR! 2021, a virtual and global conference for users and developers of the R programming language \cite{r_core_team_2021}. We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, truly inclusive and accessible experience for everyone. Most of the authors also have experience of organizing other regional and national academic conferences and events in communities of practice such as R-Ladies and The Carpentries. Here, we share the lessons learned within the past year of organizing this global useR!2021, summarized as 10 simple rules towards an inclusive conference.

@@

Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity,  geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few. Human diversity should be celebrated and respected in every way. Nonetheless, there are implicit hierarchies along these axes, and some statuses (e.g., cisgender, white, male, from the US or Europe) hold the privilege of being the default settings for which all systems --including conferences-- are consciously and unconsciously built. As a result, most people in the world fall into some category where they are part of a marginalized group --as opposed to a privileged one. A more diverse and inclusive conference  starts by recognizing that people are diverse, and that some groups face discrimination (some more than others) and might be underrepresented in your community and event \cite{timperleyHeMoanaPukepuke2020} because of these implicit hierarchies.

Think about the reasons why you want to create a more diverse conference. This may be important to you, whether you come from a privileged background or not. Along this paper, Do not treat diversity as a checklist, do not use people from minoritized groups for image purposes, and do not act as a benefactor or savior but as a genuine collaborator.

@@

Think about the groups that are usually unsupported or discriminated against in your community or event, and about the meaning of "diverse" in your context. Imagine the result you would like to see if you succeeded in organizing a diverse conference according to your vision. Would this translate into an even gender distribution in your speakers? Would it be representation of all races -especially Black people-- among the head organizers, speakers, and attendees? Would it be having LGBTQIA+ friendly-spaces or community participation from key geographic regions? This vision should guide and help hold the organizing team accountable along the way.

@@

@@

@@

Gathering a diverse team will only work if there is real inclusion. Disabled people often say: "Nothing about us without us"; the same holds for other dimensions of diversity. This means that the actual life experiences, expertise, and insights from people in marginalized groups are not replaceable with good intentions from people outside these groups. A truly inclusive and welcoming space is one in which everyone in the team is invited and allowed to bring their experience to bear. So, if you plan to organize something related to any community, region, or specific dimension of diversity, involve people from these groups.

Creating and maintaining such a team and space may seem more challenging than working in homogeneous teams, but the positive outcomes far outweigh any inconveniences. 
Having diverse people in decision-making positions will affect positively all the other aspects of your conference because all the processes will benefit from their input, expertise, and distinct perspectives \cite{hongGroupsDiverseProblem2004}. On the other hand, a diverse team plays an important role at creating a welcoming space because seeing people with similar life experiences occupy public spaces, positions of power, and breaking negative stereotypes (i.e., representation) matters (see Rule \ref{rule_inclusion}). %Representation is an important aspect of diversity and inclusion: seeing people like you, from your country, that speak your language is one of the best way to feel you belong

Most advice when working with a diverse team should apply to every team, but is especially true when working with people from marginalized groups. First, delegate. Don't expect self-nomination and voting to work as mechanisms to counteract systemic inequalities. Nominate directly and offer coordinating positions, even if this means stepping down. Offer mentorship and guidance if you encounter cases of impostor syndrome, doubts about the use of English in communication \ref{rule_language}, or others. Break the expectations about leadership as a lonely task and create smaller, co-led groups, where everyone finds their preferred tasks and gets to take leadership. Splitting the workload and responsibilities should not be done by putting care-taking labors --community building, meeting organization and note taking, conversations with potential partners-- on the hands of women and other minoritized groups, while men take the lead in stereotypical highly-valued tasks (see Rule \ref{rule_spotlight}).

Words matter. In addition to excluding derogatory or discriminatory language (remember that the code of conduct applies to you too, Rule \ref{rule_inclusion}), make the effort to teach yourself the vocabulary and the best ways to communicate to account for every culture and situation. Do not expect people to teach you --it's not their role-- and accept any feedback without being defensive.

And most importantly, take care of your team! Check on them regularly and make sure everyone is comfortable. This represents a great amount of work and you'll need to support each other in the long run.

@@

\section{Have a strong online component of the conference} 
\label{rule_online}

In-person interaction is priceless; however, it is more expensive for some, and even unattainable for others. This inaccessibility is particularly true for global conferences that usually take place in high-income countries, making it financially demanding for international participants and often impossible to attend due to immigration requirements \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019}. Online conferences are more inclusive: they do not need a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities. This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinsonJournalMedicine20202021}. Furthermore, an online format is more environmentally friendly since it eliminates travel-related emissions \cite{sarabipourChangingScientificMeetings2021,ninerBetterWhomLeveling2021, gattrellComparisonCarbonCosts2021}.

@@

Most importantly, accessibility practices are not afterthoughts that can be dealt with at the last minute. 
They require time and early decision-making \cite{irishIncreasingParticipationUsing2020}. 
Conversely, inaccessible decisions are hard to course-correct, e.g., when finding out too late that a venue is inaccessible for wheelchairs in an in-person conference. 
On the other hand, it is likely that financial constraints may not allow you to make your conference completely accessible and there will be choices to make (e.g. choosing between captions and interpreters). 
Including people that are usually disabled by standard practices in conferences in the organizing team would allow them to take part in the decisions from the beginning, spot right away the inaccessible practices that need to improve (see \textbf{Rule \ref{rule_organizing_team}}), and help make the right choices.

@@

@@

The ten rules stated here can be adapted depending on the conference format and settings.
When organizing useR!2021, we engaged in most of the practices mentioned here, and learned others along the way, so we share them here as part of our learning process. 
We organized useR! during a global pandemic, and as a team, this was a challenging journey. 
Very early on, we tried to recognize that our resources were limited (both in time of the organizing team and financially). 
We had to let go of certain aspects that were important to many of us. 
From that experience, this is our last message: Your conference may not become perfectly inclusive and accessible, but the changes you make will make a difference. 
If more conferences and domains apply these rules, the process will get more streamlined, straightforward, and mainstream to adapt with minimal overhead.
And you will make a change towards healthier, stronger, and more inclusive communities.
